Continuous Deployment and Schema Evolution \\in SQL Databases

### 2015-01-29 paper.tex
Small fixes seciton IV-VII

\author{
  \IEEEauthorblockN{
    Michael de Jong\IEEEauthorrefmark{1},
    Arie van Deursen\IEEEauthorrefmark{2}
  }
  \IEEEauthorblockA{
    Delft University of Technology,\\
    Delft, Netherlands\\ \\
    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}\\
    Email: Arie.vanDeursen@tudelft.nl\IEEEauthorrefmark{2}\\ \\
    \today{}
  }
}

@@

\section{State of the art} %(0.7)

@@

\subsection{Current tooling} 
Since evolving the schema of a database and coordinating the deployment of software are both complex tasks, we looked for tooling which could help automate this for software and release engineers.

@@

\subsection{Solution direction}

@@

\section{Implementation status} %(0.5)

\subsection{Current status}

We have started implementing a new tool --- QuantumDB~\cite{QuantumDB} --- which following this approach. The aim is to make \textit{Continuous Deployment} more attainable for software and release engineers with respect to applications relying on SQL databases. Although not yet ready for production environments, we have implemented enough to perform a quick evaluation in an small experiment.

@@

QuantumDB currently consists of two parts: 
\begin{itemize}
  \item{\textbf{core} is the program responsible for performing the actual migration tasks on the database. It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent degrading the performance of the database.}
  \item{\textbf{driver} is a Java JDBC driver which can wrap another JDBC driver. When used, the software engineers can specify a database schema version on which that version of their application operates. The driver then rewrites SQL queries sent to it by the application, and passes them on to the wrapped database JDBC driver which in turn sends them to the database. When rewriting queries, it replaces table names with the names of the associated (ghost) table names where applicable.}
\end{itemize}

@@

\section{Initial measurements} %(0.5)

We feel it is important to validate our work at an early stage. Although the development of QuantumDB is still ongoing, we setup a small experiment to verify that the concept and our implementation of it works.

\subsection{Setup}
We have installed PostgreSQL 9.3 and a small test application --- made available on GitHub~\cite{QuantumDB-RelEng-Demo} --- on a server with a 4-core Intel Core i7-3770, a 1TB hard disk, and 16GB memory. The test application inserts 50 million records into a single table called ``users'' and starts a simple demo application with two threads which continuously query and manipulate the data in that table using SELECT, INSERT, UPDATE, and DELETE statements. It then uses one of two methods to add a non nullable column to the ``users'' table. During this process we keep track of the duration of the queries issued by the demo application and how long each method takes to complete.

@@

The second approach uses QuantumDB to add the new column. It first creates a ghost table based on the ``users'' table with the new column. It then iteratively copies records from the ``users'' table to the ghost table in batches of 2,000 records at intervals of 50 milliseconds. With this we attempt to avoid negatively affecting the performance of the queries issued by the demo application while we are copying the records. Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which expose each demo application to a different version of the database schema (containing either the ``users'' table or the ghost table). The driver ensures that each query is rewritten in such a way that the correct table is queried for that specific version of the schema.

>>>>

\author{
  \IEEEauthorblockN{
    Michael de Jong\IEEEauthorrefmark{1},
    Arie van Deursen\IEEEauthorrefmark{2}
  }
  \IEEEauthorblockA{
    Delft University of Technology,\\
    Delft, The Netherlands\\ \\
    Email: M.deJong-2@student.tudelft.nl\IEEEauthorrefmark{1}\\
    Email: Arie.vanDeursen@tudelft.nl\IEEEauthorrefmark{2}\\ \\
    \today{}
  }
}

@@

\section{State of the Art} %(0.7)

@@

\subsection{Current Tooling} 
Since evolving the schema of a database and coordinating the deployment of software are both complex tasks, we looked for tooling which could help automate this for software and release engineers.

@@

\subsection{Solution Direction}

@@

\section{Implementation Status} %(0.5)

\subsection{Current Status}

We are in the process of implementing and experimenting with the approach just described via a new tool that we have named QuantumDB~\cite{QuantumDB}.
The aim is to make \textit{Continuous Deployment} more attainable for software and release engineers for applications relying on SQL databases. Although not yet ready for production environments, we have implemented enough to perform a quick evaluation in a small experiment.

@@

QuantumDB currently consists of two parts: 
\begin{itemize}
  \item{\textbf{qdb:core} is the program responsible for performing the actual migration tasks on the database. It does this by creating ghost tables of tables under change, using triggers to keep the ghost tables and the original tables synchronized with each other, and a background task which copies the data from the original tables to the ghost tables in small batches to prevent degrading the performance of the database.}
  \item{\textbf{qdb:driver} is a Java JDBC driver which can wrap another JDBC driver. When used, the software engineers can specify a database schema version on which that version of their application operates. The driver then rewrites SQL queries sent to it by the application, and passes them on to the wrapped database JDBC driver which in turn sends them to the database. When rewriting queries, it replaces table names with the names of the associated (ghost) table names where applicable.}
\end{itemize}

@@

\section{Initial Measurements} %(0.5)

In order to validate our work at an early stage we have setup a small experiment to verify that the proposed approach and our implementation of it behaves as expected.

\subsection{Setup}
We have installed PostgreSQL 9.3 and a small test application --- made available on GitHub~\cite{QuantumDB-RelEng-Demo} --- on a server with a 4-core Intel Core i7-3770, a 1TB hard disk, and 16GB memory. The test application inserts 50 million records into a single table called ``users'' and starts a simple demo application with two threads which continuously query and manipulate the data in that table using SELECT, INSERT, UPDATE, and DELETE statements. It then uses one of two methods to add a non-nullable column to the ``users'' table. During this process we keep track of the duration of the queries issued by the demo application and how long each method takes to complete.

@@

The second approach uses QuantumDB to add the new column. It first creates a ghost table based on the ``users'' table with the new column. It then iteratively copies records from the ``users'' table to the ghost table in batches of 2,000 records at intervals of 50 milliseconds. With this we attempt to avoid negatively affecting the performance of the queries issued by the demo application while we are copying the records. Unlike with the previous approach, with this approach the demo applications use the QuantumDB JDBC driver which exposes each demo application to a different version of the database schema (containing either the ``users'' table or the ghost table). The driver ensures that each query is rewritten in such a way that the correct table is queried for that specific version of the schema.
